---
title: "DPX-EEG: LMM Analysis behavioural data"
author: "José C. García Alanis"
date:  "`r Sys.Date()`"
output:
html_document:
theme: lumen
highlight: pygments
css: "styler.css"
---

Created with `r R.version.string`

```{r, setup, include = FALSE}
knitr::opts_chunk$set(class.source = "customCode",
                      class.output = "customOutput",
                      class.message = "customWarning")
if (interactive()) {
  path <- dirname(rstudioapi::getSourceEditorContext()$path)
} else {
  path <- normalizePath('./')
}
setwd(path)
source('utils.R')
```

# 0) Load necessary packages:

```{r}
load.package(
  c('rjson',
    'dplyr',
    'lme4', 'lmerTest', 'r2glmm', 'emmeans',
    'performance', 'report', 'rempsyc', 'gt')
)
```

# 1) Import the data

```{r}
paths <- fromJSON(file = 'paths.json')
fpath_tables <- paste(paths$bids,
                      'derivatives',
                      'behavioural_analysis',
                      'tables',
                      sep = '/')

# load subject data
load(
  paste(
    paths$bids,
    'derivatives',
    'behavioural_analysis',
    '00_subjects.RData',
    sep = '/')
)
# load RT data
load(
  paste(
    paths$bids,
    'derivatives',
    'behavioural_analysis',
    '01_rt_data.RData',
    sep = '/')
)
```

# 2) RT data analysis

Prepare data for analysis

```{r correct-rt-data}
# only keep sex and age variables
subjects_data <- subjects_data %>%
  ungroup() %>%
  mutate(subject = factor(subject),
         sex = factor(sex),
         age = age - mean(age)) %>%
  select(subject, sex, age)

# create factors
rt <- rt %>%
  mutate(cue_probe = paste0(C, P)) %>%
  mutate(cue_probe = factor(cue_probe,
                            levels = c('AX', 'AY', 'BX', 'BY')),
         cue = factor(C, levels = c('A', 'B')),
         probe = factor(P, levels = c('Y', 'X')),
         subject = factor(subjects)) %>%
  left_join(., subjects_data, by = 'subject') %>%
  group_by(subject, cue_probe) %>%
  mutate(reaction = factor(reaction, levels = c('Correct', 'Incorrect'))) %>%
  ungroup() %>%
  select(subject, sex, age, cue, probe, cue_probe,
         trial, run, rt, reaction, pre_error)
save(rt,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_RT_model_data.RData',
       sep = '/')
)

# clean up
rm(subjects_data)
```

## 2.1) Fit linear mixed-effects model to RT data

```{r, fit behavioral LMM}
options(width = 300)
# fit LMM
rt_mod <- lmer(data = rt,
               log(rt) ~ age +
                 sex +
                 reaction * cue * probe +
                 (cue * probe | subject:reaction),
               contrasts = list(sex = 'contr.sum',
                                reaction = 'contr.sum',
                                cue = 'contr.sum',
                                probe = 'contr.sum')
)
save(rt_mod,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_RT_mod.RData',
       sep = '/')
)

# save report
rt_mod_table <- report_table(rt_mod,
                             include_effectsize = FALSE,
                             ci = 0.99); rt_mod_table
save(rt_mod_table,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_RT_mod_report.RData',
       sep = '/')
)
rm(rt_mod_table)
```

## 2.2) Get anova table and model performance statistics (e.g., R-squared measures)

```{r, model-performance}
# get anova table
aov_rt_mod <- anova(rt_mod); aov_rt_mod
aov_rt_mod %>%
apa(title = "ANOVA table RT-model", stub = T) %>%
        fmt_number(columns = 2:7, decimals = 4) %>%
        gtsave(paste(fpath_tables,
                     'anova_RT_mod.html',
                     sep = '/'),
               inline_css = TRUE
        )
save(aov_rt_mod,
     file = paste(paths$bids,
                  'derivatives',
                  'behavioural_analysis',
                  '02_anova_RT_mod.RData',
                  sep = '/'
     )
)
```

```{r}
# compute semi-partial R2 (effect size) for each predictor by hand
eff_sizes_rt_mod <-
  ((aov_rt_mod$`NumDF` / aov_rt_mod$`DenDF`) * aov_rt_mod$`F value`) /
    (1 + ((aov_rt_mod$`NumDF` / aov_rt_mod$`DenDF`) * aov_rt_mod$`F value`))
names(eff_sizes_rt_mod) <- row.names(aov_rt_mod); eff_sizes_rt_mod
save(eff_sizes_rt_mod,
     file = paste(paths$bids,
                  'derivatives',
                  'behavioural_analysis',
                  '02_eff_sizes_RT_mod.RData',
                  sep = '/')
)

# # compute semi-partial R2 with r2glmm (computes CIs)
# rt_mod_0 <- lmer(rt_data = rt,
#                  log(rt) ~ age + sex + cue*probe + (reaction|subject))
# r2beta_rt <- r2beta(rt_mod_0, method = 'kr', partial = T, rt_data = rt); r2beta_rt
# save(r2beta_rt, file = "../rt_data/02_r2beta_RT_mod.RData")
```


```{r}
# model performance
rt_mod_rsq <- model_performance(rt_mod); rt_mod_rsq
save(rt_mod_rsq,
     file = paste(paths$bids,
                  'derivatives',
                  'behavioural_analysis',
                  "02_rsq_RT_mod.RData",
                  sep = '/')
)
rm(aov_rt_mod, eff_sizes_rt_mod, rt_mod_rsq)
```

## 2.3) Compute estimated marginal means

Compute estimated marginal means for age

```{r}
age_rt <- emmeans(rt_mod, ~ age,
                  at = list(age = c(low = -1 * sd(rt$age), high = sd(rt$age))),
                  type = 'response',
                  lmer.df = "satterthwaite",
                  lmerTest.limit = nrow(rt),
                  level = 0.99); age_rt
save(age_rt,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_age_rt_mod.RData',
       sep = '/')
)
rm(age_rt)
```

Compute estimated marginal means for sex

```{r}
sex_rt <- emmeans(rt_mod, ~ sex,
                  type = 'response',
                  lmer.df = "satterthwaite",
                  lmerTest.limit = nrow(rt),
                  level = 0.99); sex_rt
save(sex_rt,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_sex_rt_mod.RData',
       sep = '/')
)
rm(sex_rt)
```

Compute estimated marginal means for reaction

```{r}
reaction <- emmeans(rt_mod, ~ reaction,
                    type = 'response',
                    lmer.df = "satterthwaite",
                    lmerTest.limit = nrow(rt),
                    level = 0.99); reaction
save(reaction,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_reaction_RT_mod.RData',
       sep = '/')
)
rm(reaction)
```

Compute estimated marginal means for cue

```{r}
cue <- emmeans(rt_mod, ~ cue,
               type = 'response',
               lmer.df = "satterthwaite",
               lmerTest.limit = nrow(rt),
               level = 0.99); cue
save(cue,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_cue_RT_mod.RData',
       sep = '/')
)
rm(cue)
```

Compute estimated marginal means for probe

```{r}
probe <- emmeans(rt_mod, ~ probe,
                 type = 'response',
                 lmer.df = "satterthwaite",
                 lmerTest.limit = nrow(rt),
                 level = 0.99); probe
save(probe,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_probe_RT_mod.RData',
       sep = '/')
)
rm(probe)
```

Compute estimated marginal means for cue * probe interaction

```{r}
# compute estimated marginal means for cue by probe interaction
cue_by_probe <- emmeans(rt_mod, ~ cue * probe,
                        type = 'response',
                        lmer.df = "satterthwaite",
                        lmerTest.limit = nrow(rt),
                        level = 0.99); cue_by_probe
save(cue_by_probe,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_cue_by_probe_RT_mod.RData',
       sep = '/')
)
rm(cue_by_probe)
```

Compute estimated marginal means for cue * probe * reaction interaction

```{r}
# compute estimated marginal means for full interaction
interaction <- emmeans(rt_mod, ~ cue * probe * reaction,
                       type = 'response',
                       lmer.df = "satterthwaite",
                       lmerTest.limit = nrow(rt),
                       level = 0.99); interaction
save(interaction,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_interaction_RT_mod.RData',
       sep = '/')
)
rm(interaction)
```

# 3) Accuracy data

Prepare data for analysis

```{r}
# compute number of valid trials vailable per subject
total <- rt %>%
  group_by(subject, cue_probe) %>%
  mutate(n_trials = sum(!is.na(rt))) %>%
  select(subject, cue_probe, n_trials) %>%
  arrange(subject, cue_probe) %>%
  unique()

# compute number of errors per condition
errors <- rt %>%
  filter(reaction == 'Incorrect') %>%
  group_by(subject, cue_probe) %>%
  mutate(n_errors = sum(!is.na(rt))) %>%
  summarise(n_errors = mean(n_errors)) %>%
  arrange(subject, cue_probe)

# merge rt_data frames
errors <- total %>%
  left_join(., errors, by = c('subject', 'cue_probe'))
# replace missing values with zeros
errors[is.na(errors)] <- 0

# log-linear correction for error rates equal to 0.0 or 1.0
errors <- errors %>%
  mutate(error_rate_adj =(n_errors+0.5)/(n_trials+1),
         error_rate = ifelse(n_errors == 0, 0, n_errors/n_trials))

# merge with subject rt_data
subject_vars <- rt %>%
  select(subject, cue_probe, cue, probe, sex, age) %>%
  unique()
errors <- errors %>%
  left_join(., subject_vars, by = c('subject', 'cue_probe'))
# save
save(errors,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_error_data.RData',
       sep = '/')
)
```

## 3.1) Fit linear mixed-effects model to errors data

```{r, fit-accuracy-lmm}
options(width = 300)
# fit LMM
errors_mod <- lmer(data = errors,
                   log(error_rate_adj) ~ age + sex +
                     cue * probe +
                     (1 | subject),
                   contrasts = list(sex = 'contr.sum',
                                    cue = 'contr.sum',
                                    probe = 'contr.sum'))
save(errors_mod,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_errors_mod.RData',
       sep = '/')
)

# save report
errors_mod_table <- report_table(
  errors_mod,
  include_effectsize = FALSE,
  ci = 0.99); errors_mod_table
save(errors_mod_table,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_errors_mod_report.RData',
       sep='/')
)
rm(errors_mod_table)
```

## 3.2) Get anova table and model performance statistics (e.g., R-squared measures)

```{r, mod-errors-performance}
# get anova table
aov_errors_mod <- anova(errors_mod); aov_errors_mod
aov_errors_mod %>%
  apa(title = "ANOVA table error rates model", stub = T) %>%
  fmt_number(columns = 2:7, decimals = 4) %>%
  gtsave(
    paste(fpath_tables, 'anova_ER_mod.html', sep = '/'),
    inline_css = TRUE
  )
save(aov_errors_mod,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_anova_errors_mod.RData',
       sep = '/')
)
```

```{r}
# compute semi-partial R2 (effect size) for each predictor by hand
eff_sizes_errors_mod <-
  ((aov_errors_mod$`NumDF` / aov_errors_mod$`DenDF`) * aov_errors_mod$`F value`) /
    (1 + ((aov_errors_mod$`NumDF` / aov_errors_mod$`DenDF`) * aov_errors_mod$`F value`))
names(eff_sizes_errors_mod) <- row.names(aov_errors_mod); eff_sizes_errors_mod
save(eff_sizes_errors_mod,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_eff_sizes_errors_mod.RData',
       sep = '/')
)

# model performance
errors_mod_rsq <- model_performance(errors_mod); errors_mod_rsq
save(errors_mod_rsq,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_rsq_errors_mod.RData',
       sep = '/')
)
rm(aov_errors_mod, eff_sizes_errors_mod, errors_mod_rsq)
```

## 3.3) Compute estimated marginal means

Compute estimated marginal means for errors by participants' age

```{r}
age_error <- emmeans(errors_mod, ~ age,
                     at = list(age = c(low = -1 * sd(errors$age), high = sd(errors$age))),
                     type = 'response',
                     lmer.df = "satterthwaite",
                     lmerTest.limit = nrow(errors),
                     level = 0.99); age_error
save(age_error,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_age_errors_mod.RData',
       sep = '/')
)
rm(age_error)
```

Compute estimated marginal means for errors by participants' sex

```{r}
sex_error <- emmeans(errors_mod, ~ sex,
                     type = 'response',
                     lmer.df = "satterthwaite",
                     lmerTest.limit = nrow(errors),
                     level = 0.99); sex_error
save(sex_error,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_sex_errors_mod.RData',
       sep = '/')
)
rm(sex_error)
```

Compute estimated marginal means for errors by cue stimulus

```{r}
cue_error <- emmeans(errors_mod, ~ cue,
                     type = 'response',
                     lmer.df = "satterthwaite",
                     lmerTest.limit = nrow(errors),
                     level = 0.99); cue_error
save(cue_error,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_cue_error_mod.RData',
       sep = '/')
)
rm(cue_error)
```

Compute estimated marginal means for errors by probe stimulus

```{r}
probe_error <- emmeans(errors_mod, ~ probe,
                       type = 'response',
                       lmer.df = "satterthwaite",
                       lmerTest.limit = nrow(errors),
                       level = 0.99); probe_error
save(probe_error,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '02_probe_error_mod.RData',
       sep = '/')
)
rm(probe_error)
```

Compute estimated marginal means for errors by cue * probe interaction

```{r}
cue_by_probe_errors <- emmeans(errors_mod, ~ cue * probe,
                               type = 'response',
                               lmer.df = "satterthwaite",
                               lmerTest.limit = nrow(errors),
                               level = 0.99); cue_by_probe_errors
save(cue_by_probe_errors,
     file = paste(
       paths$bids,
       'derivatives',
       'behavioural_analysis',
       '/02_cue_by_probe_error_mod.RData',
       sep = '/')
)
rm(cue_by_probe_errors)
```

# 4) Compute A-cue bias
```{r}
a_bias <- errors %>%
  mutate(correct_rate = 1 - error_rate_adj) %>%
  filter(cue_probe == 'AX' | cue_probe == 'AY') %>%
  group_by(subject) %>%
  mutate(a_bias = ifelse(cue_probe == 'AX',
                         0.5 * (qnorm(correct_rate) + qnorm(lead(error_rate_adj))),
                         NA)) %>%
  select(subject, a_bias) %>%
  filter(!is.na(a_bias)) %>%
  ungroup()
```

# 4) Compute d' context
```{r}
d_context <- errors %>%
  mutate(correct_rate = 1 - error_rate_adj) %>%
  filter(cue_probe == 'AX' | cue_probe == 'BX') %>%
  group_by(subject) %>%
  mutate(d_context = ifelse(cue_probe == 'AX',
                            (qnorm(correct_rate) + qnorm(lead(error_rate_adj))),
                            NA)) %>%
  select(subject, d_context) %>%
  filter(!is.na(d_context)) %>%
  ungroup()
```

```{r}
# test correlation of performance indices
perform_index <- left_join(a_bias, d_context, by = 'subject')
correlation::correlation(data = perform_index, ci = .99)

# center indices around zero
a_bias <- a_bias %>%
  mutate(a_bias = a_bias - mean(a_bias))

d_context <- d_context %>%
  mutate(d_context = d_context - mean(d_context))

# save performance indices
write.table(d_context,
            file = paste(
              paths$bids,
              'derivatives',
              'behavioural_analysis',
              '02_d_context.tsv',
              sep = '/'),
            sep = '\t')
write.table(a_bias,
            file = paste(
              paths$bids,
              'derivatives',
              'behavioural_analysis',
              '02_a_bias.tsv',
              sep = '/'),
            sep = '\t')
```
